---
title: "02-tidy.Rmd"
author: "Jenny Bryan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    keep_md: TRUE
---

```{r include = FALSE}
## stores path to data directory into object data_dir
source("00-set-data-dir.R")
```


Recall that an important aspect of "writing data for computers" is to make your data __tidy__ (see White et al and Wickham in the Resources). There's an emerging consensus on key features of __tidy__ data:

  * Each column is a variable
  * Each row is an observation

But unfortunately, __untidy__ data abounds. In fact, we often inflict it on ourselves, because untidy formats are more attractive for data entry or examination. So how do you make __untidy__ data __tidy__?

## Import untidy Lord of the Rings data

We will now import the untidy data that was presented in the three film-specific word count tables from [the intro](01-intro.md).

I assume that data can be found as three plain text, delimited files, one for each `Film`. How to liberate data from spreadsheets or tables in word processing documents has been covered in *[TODO] this other lesson ... LINK*. We bring the data into data frames, one per `Film`, and do a little inspection. *Note: Data Carpentry stores data in the top level `data` directory; the object `data_dir` holds the path to that directory. You may have these files in your current working directory, which affects the commands below.*

```{r}
fship <- read.csv(file.path(data_dir, "The_Fellowship_Of_The_Ring.csv"))
ttow <- read.csv(file.path(data_dir, "The_Two_Towers.csv"))
rking <- read.csv(file.path(data_dir, "The_Return_Of_The_King.csv")) 
## if the csv files are in current working directory, import like so:
## fship <- read.csv("The_Fellowship_Of_The_Ring.csv")
rking
str(fship)
```

## Collect untidy Lord of the Rings data into a single data frame

We now have one data frame per `Film`, each with a common set of 4 variables. Step one in tidying this data is to glue them together into one data frame, stacking them up row wise. I'll call this row-binding and use the base function `rbind()`.

```{r}
lotr_untidy <- rbind(fship, ttow, rking)
str(lotr_untidy)
lotr_untidy
```

## Tidy the untidy Lord of the Rings data

We are still violating one of the fundamental principles of __tidy data__. "Word count" is a fundamental variable in our dataset and it's currently spread out over two variables, `Female` and `Male`. Conceptually, we need to gather up the word counts into a single variable and create a new variable, `Gender`, to track whether each count refers to females or males. We use a function from the `tidyr` package to do this.
```{r}
## uncomment and submit this line to install the tidyr package
## install.packages("tidyr")
library(tidyr)
lotr_tidy <-
  gather(lotr_untidy, key = 'Gender', value = 'Words', Female, Male)
lotr_tidy
```

Tidy data ... mission accomplished!

To explain our call to `gather()` above: we gathered the *values* in variables `Female` and `Male` together into a single new variable `Words`. A second new variable `Gender` serves as a *key*, explaining whether each of the values in `Words` refers to `Female` or `Male`. All other variables, such as `Film`, remain unchanged and are simply replicated as needed. Read the documentation for `gather()` for more information and examples.

## Write the tidy data to a delimited file

Now we write this multi-film, tidy dataset to file for use in various downstream scripts for further analysis and visualization. This would make an excellent file to share on the web with others, providing a tool-agnostic, ready-to-analyze entry point for anyone wishing to play with this data.

```{r}
write.table(lotr_tidy, file = file.path(data_dir, "lotr_tidy.csv"),
            quote = FALSE, sep = ",", row.names = FALSE)
```

You can inspect this delimited file here: [lotr_tidy.csv](../../data/tidy-data/lotr_tidy.csv).

## Exercises

The word count data is given in these two __untidy__ and gender-specific files:

  * [Female.csv](../../data/tidy-data/Female.csv)
  * [Male.csv](../../data/tidy-data/Male.csv)
  
Write an R script that reads them in and writes a single tidy data frame to file. Literally, reproduce the `lotr_tidy` data frame and the `lotr_tidy.csv` data file from above.

Write R code to compute the total number of words spoken by each `Race` across the entire trilogy. Do it two ways:

  * Using `Film`-specific or `Gender`-specific, untidy data frames as the input data.
  * Using the `lotr_tidy` data frame as input.

Reflect on the process of writing this code and on the code itself. Which is easier to write? Easier to read?

Write R code to compute the total number of words spoken in each `Film`. Do this by copying and modifying your own code for totalling words by `Race`. Which approach is easier to modify and repurpose -- the one based on multiple, untidy data frames or the tidy data?

## Take home message

It is untidy to have have data parcelled out across different files or data frames. We used `rbind()` above to combine `Film`-specific data frames into one large data frame.

It is untidy to have a conceptual variable, e.g. "word count", spread across multiple variables, such as word counts for males and word counts for females. We used the `gather()` function from the `tidyr` package to stack up all the word counts into a single variable, create a new variable to convey male vs. female, and  do the replication needed for the other variables.

Many data analytic projects will benefit from a script that marshals data from different files, tidies the data, and writes a clean result to file for further analysis.

Watch out for how __untidy__ data seduces you into working with it more than you should:

  * Data optimized for consumption by human eyeballs *is* attractive, so it's hard to remember it's suboptimal for computation. How can something that looks so pretty be so wrong?
  * Tidy data often has lots of repetition, which triggers hand-wringing about efficiency and aesthetics. Until you can document a performance problem, keep calm and tidy on.

### Where to next?

In the [optional bonus content](03-tidy-bonus-content.md), I show how to tidy this data using only base R functions. At the other extreme, I also show how to tidy with add-on packages that are capable of more advanced data manipulations.

### Resources

  * [Bad Data Handbook](http://shop.oreilly.com/product/0636920024422.do) by By Q. Ethan McCallum, published by O'Reilly.
    - Chapter 3: Data Intended for Human Consumption, Not Machine Consumption by Paul Murrell.
  * Nine simple ways to make it easier to (re)use your data by EP White, E Baldridge, ZT Brym, KJ Locey, DJ McGlinn, SR Supp. *Ideas in Ecology and Evolution* 6(2): 1–10, 2013. doi:10.4033/iee.2013.6b.6.f <http://library.queensu.ca/ojs/index.php/IEE/article/view/4608>
  * Tidy data by Hadley Wickham. Journal of Statistical Software. Vol. 59, Issue 10, Sep 2014. <http://www.jstatsoft.org/v59/i10>
    - [`tidyr`](https://github.com/hadley/tidyr), an R package to tidy data.
    - R packages by the same author that do heavier lifting in the data reshaping and aggregation department include [`reshape2`](https://github.com/hadley/reshape), [`plyr`](https://github.com/hadley/plyr) and [`dplyr`](https://github.com/hadley/dplyr).
