---
title: "03-tidy-bonus-content.Rmd"
author: "Jenny Bryan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    keep_md: TRUE
---

```{r include = FALSE}
## stores path to data directory into object data_dir
source("00-set-data-dir.R")
```

This material is not part of the lesson. But it may be helpful when tidying data in real life.

## More about `rbind()`ing data frames

In [the main tidy lesson](02-tidy.md), the first step was to stack up `Film`-specific data frames row-wise. First I redo that, then present alternative methods for `rbind()`ing data frames.

```{r}
fship <- read.csv(file.path(data_dir, "The_Fellowship_Of_The_Ring.csv"))
ttow <- read.csv(file.path(data_dir, "The_Two_Towers.csv"))
rking <- read.csv(file.path(data_dir, "The_Return_Of_The_King.csv"))  
lotr_untidy <- rbind(fship, ttow, rking)
```

Note that the 3 underlying data frames are hard-wired into this command: `rbind(fship, ttow, rking)`. In real life, this can become very cumbersome and even impossible, if there is a large number of data frames that need to be combined. Things get challenging for two reasons:

  * listing the data frames explicitly can be a drag and error prone
  * the actual `rbind()`ing can take lots of memory and time
  
### Memory efficient row-binding

The `dplyr` package offers the function `rbind_list()` as an efficient substitute for the base `rbind()`.

```{r}
suppressPackageStartupMessages(library(dplyr))
lotr_untidy_2 <- rbind_list(fship, ttow, rking)
```

We get a warning about row-binding data frames with factors that don't have the same levels (here, the `Film` factor). The base function `rbind()` handles this by taking the union of factor levels, whereas `dplyr::rbind_list()` converts the affected factor to character.

```{r}
str(lotr_untidy)
str(lotr_untidy_2)
```

It is easy to make `Film` back into a factor:

```{r}
lotr_untidy_2$Film <-
  factor(lotr_untidy_2$Film,
         levels = c("The Fellowship Of The Ring", "The Two Towers",
                    "The Return Of The King"))
```

The advantage of `dplyr::rbind_list()` over base `rbind()` will become apparent when the data frames you are row-binding are large and/or numerous. 

### Row-binding a list of data frames

Frequently the data frames destined for row-binding are collected together in a list. Here are several ways to proceed, starting with the most primitive.

To prepare, we collect the `Film`-specific data frames into a single list.

```{r}
lotr_files <- file.path(data_dir, c("The_Fellowship_Of_The_Ring.csv",
                                  "The_Two_Towers.csv",
                                  "The_Return_Of_The_King.csv"))
lotr_list <- lapply(lotr_files, read.csv)
str(lotr_list)
```

FYI, `lapply()` is one of the base R functions for data aggregation; it iteratively applies a function to each element of a vector.

#### Base R, brute force

We can use `rbind()` as before and give each data frame explicitly by specifying the 1st, 2nd, and 3rd elements of the list.

```{r}
lotr_untidy_3 <- rbind(lotr_list[[1]], lotr_list[[2]], lotr_list[[3]])
str(lotr_untidy_3)
```

As you can imagine, this *really* does not scale well. What if there were 20 data frames in this list?!? Or 200?

#### Base R, `do.call()`

The arcane-sounding function `do.call()` "constructs and executes a function call from a name of a function and a list of arguments to be passed to it". Although the use of `do.call()` is not limited to `rbind()`, this is perhaps the most common use case:

```{r}
lotr_untidy_4 <- do.call(rbind, lotr_list)
str(lotr_untidy_4)
```

This is a huge improvement over the brute force solution, because the individual data frames are no longer explicitly listed, one by one.

#### `dplyr::rbind_all()`

The `dplyr` package offers a memory-efficient solution for row-binding a list of data.frames, namely `rbind_all()`.

```{r}
lotr_untidy_5 <- rbind_all(lotr_list)
str(lotr_untidy_5)
```

We get the same warning as before about unequal factor levels for `Film`; resolve as shown above, if you want `Film` to be factor vs character.

The `rbind_all()` function from `dplyr` probably represents the best all around solution, because it addresses both pain points at once: it is memory efficient and it can operate on a list of data frames.

#### Other options

Other options for row binding data frames (and more) include the [`rbindlist()` function](http://stackoverflow.com/questions/15673550/why-is-rbindlist-better-than-rbind) from the [`data.table`](https://github.com/Rdatatable/data.table/) package and the `rbind.fill()` function from the `plyr` package. [This comparison](http://rcrastinate.blogspot.ca/2013/05/the-rbinding-race-for-vs-docall-vs.html) of row binding methods is informative, though it would be good to expand to include `data.table::rbindlist()` and `dplyr::rbind_all()`.

## More about gathering variables

In [the main tidy lesson](02-tidy.md), the second step was to gather the word counts stored as separate variables for `Females` and `Males` and stack them up to make two new variables: `Words` and `Gender`.

We start with the untidy data frame that results from any of the row-binding methods above.

```{r}
lotr_untidy
```

Here we repeat the `tidyr::gather()` approch from the main lesson, but also present alternatives that use no add-on packages at all and that use the more powerful `reshape2` package.

#### Base R, brute force

It is entirely possible to reshape data "by hand". Here we exploit R's recycling behavior to replicate the variables `Film` and `Race`. We create the new variable `Words` by concatenating `Female` and `Male` and we create a new factor `Gender`.

```{r}
lotr_tidy <-
  with(lotr_untidy,
       data.frame(Film = Film,
                  Race = Race,
                  Words = c(Female, Male),
                  Gender =rep(c("Female", "Male"), each = nrow(lotr_untidy))))
lotr_tidy
```

#### Base R, `stack()`

I do not consider `stack()` useful in real life, given all the alternatives. Including only for completeness.

```{r}
lotr_tidy_2 <-
  with(lotr_untidy,
       data.frame(Film = Film,
                  Race = Race,
                  stack(lotr_untidy, c(Female, Male))))
names(lotr_tidy_2) <- c('Film', 'Race', 'Words', 'Gender')
lotr_tidy_2
```

#### `tidyr::gather()`

This is repeated content from the main lesson.

```{r}
library(tidyr)
lotr_tidy_3 <-
  gather(lotr_untidy, key = 'Gender', value = 'Words', Female, Male)
lotr_tidy_3
```

#### `reshape2::melt()`

The `reshape2` package is more powerful than `tidyr` but also harder to use and often overkill. But some reshaping tasks are beyond the capabilities of `tidyr`, so its good to know `reshape2` is there when you need it.

```{r}
library(reshape2)
lotr_tidy_4 <-
  melt(lotr_untidy, measure.vars = c("Female", "Male"), value.name = 'Words')
lotr_tidy_4
```

In `reshape2` jargon, we want to `melt()` the untidy LOTR data. Under the hood, since we are `melt()`ing a data frame, the function `melt.data.frame()` is what's  actually used; read [the documentation](http://www.rdocumentation.org/packages/reshape2/functions/melt.data.frame). The first argument `data =` specifies the data frame to work on. The `measure.vars =` argument specifies the variables that should be gathered together to make a new variable -- here `Female` and `Male` word counts. The remaining variables -- `Film` and `Race` -- are assumed to be `id.vars` and will be replicated as necessary. Finally, if you want to name your new variable, indicate that via the `value.name =` argument.

Since `melt()` "will assume factor and character variables are id variables, and all others are measured," we could have just called it like so, but this seemed too magical to be useful for teaching!

```{r eval = FALSE}
melt(lotr_untidy, value.name = 'Words')
```

## Resources

`dplyr` package: on [GitHub](https://github.com/hadley/dplyr) | an [introduction vignette](http://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html)

`data.table` package: on [GitHub](https://github.com/Rdatatable/data.table/)

`tidyr` package: on [GitHub](https://github.com/hadley/tidyr),

`reshape2` package: on [GitHub](https://github.com/hadley/reshape)